---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# explan

<!-- badges: start -->
<!-- badges: end -->

The goal of `{explan}` is to help researchers design psychology experiments. Currently this package provides an implementation of Permuted Subblock Randomization (PSR), a restricted randomization approach that improves power for within-participant study designs. See [Liang & Barr (2024)](https://osf.io/preprints/psyarxiv/4ums9) for details.

```
@article{liang_barr_2024,
 title={Better power by design: Permuted-Subblock Randomization boosts power in repeated-measures experiments},
 url={osf.io/preprints/psyarxiv/4ums9},
 DOI={10.31234/osf.io/4ums9},
 publisher={PsyArXiv},
 author={Liang, Jinghui and Barr, Dale J},
 year={2024}
}
```

The long-term goal is to provide a larger set of functions to help with counterbalancing, Latin Square designs, etc.

## Installation

You can install the development version of explan from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("dalejbarr/explan")
```

## Implementing Permuted Subblock Randomization

The package includes two main functions for implementing PSR, a vector version (`psr()`) and a data frame version (`psr_stimuli()`).

The vector version, `psr()` creates a vector of condition levels conforming to PSR given a set of levels corresponding to experimental conditions, desired number of subblocks, and number of repetitions per condition.

```{r example}
library(explan)

psr(c("condition-A", "condition-B", "condition-C"),
    n_subblocks = 6, n_reps = 6)
```

To find out the number of possible subblocks for a given number of repetition, use `possible_subblocks()`:

```{r poss-subblocks}
possible_subblocks(6)
```

The data frame version, `psr_stimuli()`, is a function that makes it possible to apply PSR to a table of stimuli. As an example, the built-in table `stroop_stimuli` contains example stimuli for a Stroop experiment, which has a two-level within participant factor of congruency.

```{r stroop-stimuli}
stroop_stimuli
```

You can determine the number of replications using `dplyr::count()`:

```{r congruency-count}
stroop_stimuli |>
  dplyr::count(congruency)
```

```{r poss-subblocks-stroop}
possible_subblocks(24)
```

To randomize this stimuli for a study with four participants, you would use the following code:

```{r randomize-stroop}
stroop_four <- psr_stimuli(stroop_stimuli, "congruency",
                           n_subblocks = 12, n_part = 4)

stroop_four |>
  head(10)
```

<!-- build the README.md with devtools::build_readme() -->

<!-- TODO: mention vignette -->
